---
title: "Safely Store xcms Result Objects in a Portable Stash"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Safely Store xcms Result Object in a Portable Stash}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{SpectraStash}
    %\VignetteDepends{MsExperiment,Spectra,SpectraStash,BiocStyle,alabaster.base,fs,xcms,faahKO}
---

```{r style, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(BiocStyle)
```

# Introduction

Data objects in R can be serialized to disk in R's *rds* or *RData* format using
the base R `save()` function and re-imported using the `load()` function. This
binary file format can however not be used easily in other programming
languages. The `r Biocpkg("MsStash")` package defines a basic framework to store
and import mass spectrometry (MS) data objects in various storage formats aiming
to simplify and facilitate data exchange between software. The *xcmsStash*
package implements portable data storage formats (stashes) for result objects
from Bioconductor's `r Biocpkg("xcms")` package. Supported stash formats are,
next to storage in simple plain text files, also Bioconductor's *alabaster*
format defined in the `r Biocpkg("alabaster.base")` and related packages.

# Installation

The package can be installed with the *BiocManager* package. To install
*BiocManager* use `install.packages("BiocManager")` and, after that, install
this package along with all required dependencies using:

```{r install, eval = FALSE}
#' Install xcmsStash along with all required dependencies
BiocManager::install("xcmsStash")
```

# A stash for *xcms* result objects

*xcms* result objects can be saved and restored through the `saveMsObject()` and
`readMsObject()` functions into (or from) MS data stashes. At present,
`XcmsExperiment` and `XcmsExperimentHdf5` objects can be saved in the following
stash formats:

- *Alabaster*-based format: storage of MS data and *xcms* results using
  Bioconductor's `r Biocpkg("alabaster.base")` framework saving files in HDF5
  and JSON format. This stash format can be selected and configured by passing
  an `AlabasterParam` to the `saveMsObject()` or `readMsObject()`
  function. *alabaster*-based stashes also fully support the `saveObject()` and
  `readObject()` functions from *alabaster.base*.
- Text file-based format: storage of the MS data and preprocessing results in
  plain text files. This stash format can be selected and configured by passing
  a `PlainTextParam` to the `saveMsObject()` and `readMsObject()` function.

See also the vignette from the `r Biocpkg("MsStash")` package for details on the
formats and the documentation and vignettes of the related
`r Biocpkg("SpectraStash")` and `r Biocpkg("MsExperimentStash")` packages for
information on storage of the result object's MS data.

Below we load an example `XcmsExperiment` result object provided by the *xcms*
package.

```{r load-result, message = FALSE}
library(xcms)
library(MsExperiment)
library(xcmsStash)

xmse <- loadXcmsData("xmse")
xmse
```

The `XcmsExperiment` thus contains the full results from a *xcms*-based LC-MS
data preprocessing, chromatographic peaks, adjusted retention times and LC-MS
features. We next store this object to a *xcmsStash* using the `saveMsObject()`
function with an `AlabasterParam` parameter object to choose and configure the
*alabaster*-based stash format. The location of the stash can be set
with the `path` argument of this parameter object. For the present example we
save the stash to a temporary folder.

```{r save-alabaster}
#' Define the location of the stash
d <- file.path(tempfile(), "xmse_stash")

#' Configure the format and location
ap <- AlabasterParam(d)

#' Save the `XcmsExperiment` object to the stash
saveMsObject(xmse, ap)
```

`saveMsObject()` was storing, next to the results present in the
`XcmsExperiment` object also all other experiment-related information from the
MS data containers `MsExperiment` and `Spectra`, that the object contains.
The content of the thus created stash folder is:

```{r show-directory}
library(fs)
dir_tree(d)
```

An `XcmsExperiment` contains, next to the results from the *xcms* preprocessing,
also the MS data of the experiment along with the experiment's sample
information. All the experiment-relevant information is contained in a
`MsExperiment` object (which the `XcmsExperiment` extends), with the actual MS
data of the experiment being represented by a `Spectra` object. Calling
`saveMsObject()` on a `XcmsExperiment` triggers the serialization mechanism of
all these contained or extended MS data objects. In alabaster format each slot
of an object is then stored to a sub-directory within the main *xcmsStash*
directory. Folders *chrom_peak_data*, *chrom_peaks*, *feature_definitions* in
the stash folder above contain the *xcms* preprocessing results, and the
*xcms_experiment_process_history.json* file provides the full data processing
history (all parameter objects from the object's `processHistory()` serialized
in JSON format). All other folders contain the remaining MS experiment-relevant
information in their respective *sub-stashes*. The *spectra* folder contains for
example the experiment's MS data as a *SpectraStash* (defined in the
`r Biocpkg("SpectraStash")` package).

An `XcmsExperiment` can be restored from an *xcmsStash* with the
`readMsObject()` function along with the parameter object specifying the format
of the stash. To load the result object from the alabaster-format stash created
above we use the `AlabasterParam()` parameter object providing the file path to
the stash as argument. In addition we need to define the expected result
(return) object with the first parameter to `readMsObject()` (in our case an
`XcmsExperiment()` object).

```{r load-alabaster-stash}
#' Restore the result object
res <- readMsObject(XcmsExperiment(), AlabasterParam(d))
res
```

As mentioned above, *xcmsStash* folders are organized in a modular way with the
different components being stored in sub-folders (*sub-stashes*). This allows to
load only selected parts from the object instead of the full object. We can for
example read only the MS experiment information from the stash, omitting the
*xcms* preprocessing results.

```{r load-ms-experiment}
#' Load the MS experiment data from the stash
mse <- readMsObject(MsExperiment(), AlabasterParam(d))
mse
```

Or even just the `Spectra` object representing the MS data of the experiment
from the *spectra* sub-folder of the stash directory:

```{r load-spectra}
#' Load the Spectra object from the stash
library(Spectra)
sps <- readMsObject(Spectra(), AlabasterParam(file.path(d, "spectra")))
sps
```

Depending on the data backend the `Spectra` object of an `XcmsExperiment` uses,
only references to the actual MS data might be stored within the *xcmsStash*. As
shown by the print output above, the loaded `Spectra` object uses a
`MsBackendMzR` backend to represent the MS data. This type of backend only loads
general metadata information into memory and reads the actual MS data only
on-demand from the original MS data files. We can use the
`dataStorageBasePath()` function to get the location of these original MS data
files for this `Spectra` object.

```{r ms-data-location}
#' Get the location of the original MS data files
dataStorageBasePath(sps)
```

The MS data files are thus only referenced from the `Spectra` object but not
part of the stash folder. If the original data files are moved to another
location or if the *xcmsStash* folder is moved to a different computer, the
location of these MS data files needs to be provided with the `spectraPath`
parameter of the `readMsObject()` function to allow restoring the object from
the stash (see documentation and vignette of the `r Biocpkg("SpectraStash")`
package for more information).

Alternatively, it is possible to store also the MS data files **into** the stash
folder by passing `consolidate = TRUE` to the `saveMsObject()` call:

```{r consolidate-stash}
#' Store the result object consolidating the full data into the stash
d2 <- file.path(tempdir(), "consolidated_xcms_stash")
saveMsObject(xmse, AlabasterParam(d2), consolidate = TRUE)
```

Now the stash directory contains also the original MS data files, in that
particular example stored within sub-folders *KO* and *WT* within the *spectra*
sub-folder:

```{r consolidate-directory-content}
#' List the directory content of the consolidated *xcmsStash*
dir_tree(d2)
```

While this allows generating self-contained portable stashes, their file size
will be much larger, depending on the number and size of the original MS data
files.

Finally, the *xcmsStash* package implements also the `saveObject()` and
`readObject()` methods from the `r Biocpkg("alabaster.base")` package. Thus,
the full stash, or also selected parts of a result object can be read using
these functions. Loading below for example the `Spectra` object of the *xcms*
result object using *alabaster*'s `readObject()` function:

```{r alabaster-read}
#' Read the `Spectra` object from the *xcmsStash*
library(alabaster.base)
sps <- readObject(file.path(d2, "spectra"))
sps
```

Next to the alabaster-based stash format *xcmsStash* implements also a *plain
text file*-based format (`PlainTextParam`), that stores all data as (human
readable) tabulator delimited text files. Additional formats, such as the
possibility to save *xcms* results in mzTab-M format (through the
`r Biocpkg("RmzTabM")` package) will be added in future.

# Session information

```{r session-information}
sessionInfo()
```
